Re-examining cross-cultural similarity judgements using lexical co-occurrence

### 2023-01-19 code/4_paper/manuscript/manuscript.Rmd
fixed merge

abstract: >
 Is “cow” more closely related to “grass” or “chicken”? Speakers of different languages judge similarity in this context differently, but why? One possibility is that cultures co-varying with these languages induce variation in conceptualizations of similarity. Specifically, East Asian cultures may promote reasoning about thematic similarity, by which cow and grass are more related, whereas Western cultures may bias similarity judgements toward taxonomic relations, like cow-chicken. This difference in notions of similarity is the consensus interpretation for cross-cultural variation in this paradigm. We consider, and provide evidence for, an alternative possibility, by which notions of similarity are equivalent across contexts, but the statistics of the environment vary. On this account, similarity judgements are guided by co-occurrence in experience, and observing or hearing about cows and grass or cows and chickens more often could induce preferences for the relevant grouping, and account for apparent differences in notions of similarity across contexts.

@@

Taxonomic and thematic similarity provide a convenient entry point to broader debates about cross-linguistic and cross-cultural variation in notions of similarity. Taxonomic categorization is based on the similarity of attributes, for example, similar perceptual properties, like shared color or shape, among objects. In contrast, thematic categorization is based on causal, spatial, and temporal relationships among objects [@Markman1984]. To illustrate this point, if a participant is given the words “chicken” and “grass”, and asked which of those words are more closely related to the word “cow”, the pairing “chicken” - “cow” indicates a taxonomic match, whereas “grass” - “cow” indicates a thematic match.

@Ji2004 found that preferences for taxonomic and thematic matches differ across cultures, even when controlling for testing language and fluency. Specifically, Chinese participants made more thematic matches than taxonomic, while the converse was true for European Americans. This difference held even for Chinese participants tested in English (rather than Mandarin), suggesting that culture has an effect on the categorization strategy independent of the test language. However, the study also found an effect of language within Chinese participants from mainland China (when tested both in China and in the US): their preference for thematic categorization was significantly lower when tested in English compared to Mandarin Chinese. This suggests that test language also influences categorization strategies. Ji et al. characterize these results as demonstrating that culture (independent of language) leads to different styles of reasoning about similarity, whereas language serves as a “tuning” mechanism operating within a culturally-specific style, by activating representations corresponding to the language being used.

## Cross-cultural variation in similarity
The view that culture drives differences in categorization style has been one favored by other non-linguistic studies that use taxonomic vs thematic categorization to probe similarity judgment. When shown pictorial triads that represent words (i.e. pictures of an ant, a bee, and honey), Chinese children (9-10 years old) scored higher on thematic categorization compared to American children of the same age range, and vice versa for taxonomic categorization [@Chiu1972]. These cross-cultural differences in similarity judgment persist in novel object categorization, with Chinese participants preferring to group by family resemblance across multiple features and Americans preferring a single-feature rule [@Norenzayan2002]. Since differences in categorization style persist with novel stimuli (as opposed to previously seen words or pictures of concepts), this suggests that participants from different cultural backgrounds are utilizing different reasoning strategies when making similarity judgments. The authors attribute these differences to tendencies toward analytic processing in Western cultures, which emphasizes objects and their properties and holistic processing in East Asian cultures, which emphasizes relationships between objects and their context. According to this framework, East Asians attend more to relationships between objects and context and reason more holistically, whereas Westerners attend more to objects and their properties, and reason analytically (ibid, see also @Nisbett2003). In related work, East Asian participants show a higher level of sensitivity to context than their Western counterparts when reproducing drawings from memory [@Ji2000]; visually exploring naturalistic scenes [@Chua2005]; describing scenes [@Masuda2001]; and in explaining the causes of ambiguous behaviors [@Choi1999]. These cross-cultural differences in attention and analytic vs holistic processing are seen to map onto the cross-cultural differences in similarity judgment: because East Asian participants attend more to the context between objects and reason more holistically, it stands to reason that they would consider words that appear in the same context (i.e. thematically similar) to be more similar than taxonomically-similar words.

This view ascribes cross-cultural differences in similarity judgment tasks to variation in the conceptualization of similarity itself. Alternatively, these judgments could be shaped by cross-cultural differences in the statistics of the environment, and accordingly the content of everyday experiences. Perhaps when confronted with the triad task, participants from all cultures follow the same process for reasoning about similarity, but rely on language or culture-specific inputs to this process. For example, it may be that Vietnamese participants are more likely to have knowledge of farming practices than Americans, and may associate cows and grass more strongly (as grass is the typical feed for cows in the small-scale family farming contexts common in Vietnam). Americans, on the other hand, may have strong associations between cows and chickens as a result of encountering them together in children’s books or petting zoo contexts. If we observe a difference in categorization between, say Vietnamese and US participants, it could be that members of both groups use the same procedure (considering similarity that is influenced by both taxonomic and thematic relations), but the inputs to this procedure differ between cultures, with Vietnamese participants exposed to more instances of “cow” and “grass” occurring together, compared to US participants.

## Language as a proxy of variation in experience
While co-occurrences in experience are difficult to measure, co-occurrence in language can provide a rough proxy--and indeed, language may afford many of the “experiences” that people have with infrequently encountered items, like cows or helicopters. In this study, we use co-occurrence in language as a proxy of cultural experience. This is a generalizing assumption, and while no linguistic corpus can be expected to fully capture a culture, using this proxy as a model of culture provides a conservative test of the hypothesis that differences in the input, rather than the process, of reasoning produces cultural variation in this task. This means that if such a model does work -- if linguistic co-occurrence can predict cultural variation in this similarity task without building differences in the decision process into the model -- it is relatively strong evidence that language alone (perhaps as a more easily observed source of information on broader cultural and ecological variation) may account for these cross-cultural differences in reasoning.

While we discuss cross-cultural variability at the level of countries or larger world areas, these are not cultural monoliths. For convenience, we operationalize culture at the level of country, based on where participants were raised. It is an open question whether performance in our participant populations (of relatively young and well-educated adults) is representative of the broader country. This is especially true for societies with substantial ethnic and cultural variation such as the US. We expect that our data is likely to underestimate variation both within and between the countries we sample from.

Indeed, previous work suggests that using lexical co-occurrence as a proxy can be useful in thinking about similarity judgments. Natural language processing tasks have found that lexical co-occurrence is a good predictor for performance in similarity judgment tasks. @Griffiths2007 showed that a model that takes word-document co-occurrence as input can be used to predict word association and the effects of semantic association on a variety of linguistic tasks. Additionally, lexical semantic models that are constructed using lexical co-occurrence (as opposed to annotated relations) have been shown to perform well on predicting human judgments about similarity between word pairs that are either thematically or taxonomically related [@Rohde2006].

Our current study will be using 2 measures of lexical co-occurrence: 1) raw lexical co-occurrences and 2) cosined distances of fastText word vectors. fastText is a system that uses lexical co-occurrence information to generate a vector representing each word in its lexicon [@Mikolov2018]. fastText uses a continuous bag-of-words model, which takes into account a symmetric window of words surrounding the word in question and maximizes the log-likelihood of the probability of the word given its context. Calculating the cosine-similarity between word vectors from fastText gives us a score of how similar the contexts of these words are. Therefore, words that occur in the same context (high lexical co-occurrence) would have a larger cosine-similarity score.

It must be noted that fastText supplements the lexical co-occurrence information by simultaneously training weight vectors for each position relative to the word in question, as well as adding sub-word information (bag-of-character n-gram vectors for each character forming the words). The similarity judgment prediction given by fastText may be more higher than a model using purely lexical co-occurrence information. However, models that use only lexical co-occurrence information (with or without a simple position weight system that gives more weight to words that are closer to the word in question) also perform well in similarity prediction tasks. For this study, due to the lack of large cleaned corpora for Mandarin and Vietnamese, we are leveraging fastText pre-trained models on the languages we are interested in @Grave2018.

Systems like fastText and word2vec have been demonstrated to be good predictors for similarity judgments. @Liu2019 showed that fastText performed well as the data for a system matching responses in a knowledge base by how similar they are to a customer’s query. A study by @Jatnika2019 tested a word2vec model trained on the English wikipedia corpus against 353 word pairs labeled with similarity values based on human judgment. The word2vec model correlates moderately with human judgment. fastText has also been shown to be sensitive towards cultural effects on word meanings: @Thompson2020 applied fastText to language-specific Wikipedia corpus (among others) to generate a “semantic neighborhood” of 1010 meanings in different languages. The study showed that languages that are spoken by more similar cultures, more geographically proximate and/or historically related had more semantic neighborhoods that aligned. This is a proof of concept for our use of fastText in different languages with varying levels of relatedness.

@@

<!-- Additionally, outside of language, cross-cultural differences could come from a couple different mechanisms:  -->
In this study, we ask whether variation in lexical co-occurrence can account for the cross-context (culture and language) differences observed in how people evaluate similarity. First, we attempt to replicate @Ji2004 by running a study comparing English speakers in the US and Mandarin speakers in mainland China. Second, we evaluate whether the differences observed between English and Mandarin speakers’ similarity judgments also extend to comparisons between English and Vietnamese, by running a similar study between English speakers in the US and Vietnamese speakers in Vietnam. Vietnam is a Southeast Asian country that borders China and is historically greatly influenced by Chinese culture [@Hui2002]. Therefore, it serves as a suitable cultural context to investigate whether the claim made by @Ji2004 and previous studies, that Eastern and Western cultures have different notions of similarity, extends beyond mainland China.

We then ask to what extent similarity judgments show cross-context differences that align with co-occurrence patterns in participants’ languages. We would only see a main effect of cultural context if differing notions of similarity drive these cross-cultural differences. Meanwhile, if we do see variation in responding trial-to-trial that tracks with lexical co-occurrence statistics (as a rough proxy combining both linguistic and cultural context), this suggests that cross-cultural notions of similarity might be similar, and it is environmental statistics that guide the differences in similarity judgments. To preview our results, we find that, while we do not see overall group differences (between the East Asian contexts and the US) in similarity, we do find language-specific co-occurrence can indeed explain cross-context differences in similarity judgments. In contrast to previous accounts, by which culture induces differing conceptions of similarity (varying between taxonomic and thematic), these findings provide an alternative, and more specific, account by which language may explain these cross-cultural differences without invoking variation in notions of similarity.

>>>>

abstract: >
 Is “cow” more closely related to “grass” or “chicken”? Speakers of different languages judge similarity in this context differently, but why? One possibility is that cultures co-varying with these languages induce differences in conceptualizations of similarity. Specifically, East Asian cultures may promote reasoning about thematic similarity, by which cow and grass are more related, whereas Western cultures may bias judgements toward taxonomic relations, like cow-chicken. This difference in notions of similarity is the consensus interpretation for cross-cultural variation in this paradigm. We consider, and provide evidence for, an alternative possibility, by which notions of similarity are similar across contexts, but the statistics of the environment vary. On this account, similarity judgements are guided by co-occurrence in experience, and observing or hearing about cows and grass or cows and chickens more often could induce preferences for these groupings, and account in part for apparent differences in notions of similarity across contexts.

@@

Taxonomic and thematic similarity provide a convenient entry point to broader debates about cross-linguistic and cross-cultural variation in notions of similarity. Taxonomic categorization is based on the similarity of attributes, for example, similar perceptual properties, like shared color or shape, among objects. In contrast, thematic categorization is based on causal, spatial, and temporal relationships among objects [@Markman1984].

## Cross-cultural variation in similarity
Preferences for taxonomic and thematic similarity vary across cultures. @Chiu1972 found that Chinese children (9-10 years old) are more likely to choose thematic matches in a picture triad task (shown, e.g., images of an ant, a bee, and honey) than their American counterparts. These cross-cultural differences are also observed in novel object categorization, with Chinese participants preferring to group by family resemblance across multiple features and Americans preferring a single-feature rule [@Norenzayan2002]. 
<!-- Because differences in similarity preference persist with novel stimuli (as opposed to familiar items), this suggests that participants from different cultural backgrounds use different reasoning strategies when making similarity judgments. -->
The authors link these differences to tendencies toward analytic processing in Western cultures, which emphasizes objects and their properties, and holistic processing in East Asian cultures, which emphasizes relationships between objects and their context (see also @Nisbett2003). In related work, East Asian participants show a higher level of sensitivity to context than their Western counterparts when reproducing drawings from memory [@Ji2000]; visually exploring naturalistic scenes [@Chua2005]; describing scenes [@Masuda2001]; and in explaining the causes of ambiguous behaviors [@Choi1999].

@Ji2004 asked whether differences in analytic and holistic processing <!--they use that language, right?--> are driven by language, culture, or a combination of these. They presented Chinese and European American adults with a word triad task, and found a preference for thematic matching among Chinese participants compared to European Americans. They found that test language contributed to this difference, but did not fully explain it: Chinese participants showed a stronger thematic preference when tested in Mandarin, but still showed a thematic preference when tested in English. Ji et al. conclude that culture (independent of language) leads to different styles of reasoning about similarity, whereas language serves as a “tuning” mechanism operating within a culturally-specific style, by activating representations corresponding to the language being used.^[But note that this work did not include a manipulation to prevent participants from engaging in e.g., covert naming in a language other than the test language.]

This view ascribes cross-cultural differences in similarity judgment to variation in the conceptualization of similarity itself. Alternatively, these judgments could be shaped by cross-cultural differences in the statistics of the environment, and accordingly the content of everyday experiences. Perhaps when confronted with the triad task, participants from all cultures follow the same process for reasoning about similarity, but rely on language or culture-specific inputs to this process. <!--For example, it may be that Vietnamese participants are more likely to have knowledge of farming practices than Americans, and may associate cows and grass more strongly (as grass is the typical feed for cows in the small-scale family farming contexts common in Vietnam). Americans, on the other hand, may have strong associations between cows and chickens as a result of encountering them together in children’s books or petting zoo contexts.--> If we observe a difference in categorization between East Asian and Western participants, it could be that members of both groups use the same procedure (considering similarity that is influenced by both taxonomic and thematic relations), but the inputs to this procedure differ between cultures, with East Asian participants exposed to more instances of thematic similarity than their Western counterparts.

## Estimating variation in experience
While co-occurrences in experience are difficult to measure, co-occurrence in language can provide a rough proxy--and indeed, language may afford many of the “experiences” that people have with infrequently encountered items, like cows or helicopters. In this study, we use co-occurrence in language as a proxy of cultural experience. This is a generalizing assumption, and while no linguistic corpus can be expected to fully capture a culture, using this proxy as a model of culture provides a conservative test of the hypothesis that differences in the input, rather than the process, of reasoning produces cultural variation in this task. This means that if such a model does work -- if linguistic co-occurrence can predict cultural variation in this similarity task without building differences in the decision process into the model -- it is relatively strong evidence that language alone (perhaps as a more easily observed source of information on broader cultural and ecological variation) may account for these cross-cultural differences in reasoning.

Indeed, previous work suggests that using lexical co-occurrence as a proxy can be useful in thinking about similarity reasoning. Natural language processing tasks have found that lexical co-occurrence is a good predictor for performance in similarity judgment. @Griffiths2007 showed that a model that takes word-document co-occurrence as input can be used to predict word association and the effects of semantic association on a variety of linguistic tasks. Additionally, lexical semantic models that are constructed using lexical co-occurrence (as opposed to annotated relations) have been shown to perform well on predicting human judgments about similarity between word pairs that are thematically or taxonomically related [@Rohde2006].

Our study uses two measures of lexical co-occurrence: 1) raw lexical co-occurrence and 2) cosined distances of fastText word vectors. fastText is a system that uses lexical co-occurrence information to generate a vector representing each word in its lexicon [@Mikolov2018].

<!-- MOVE TO METHODS? fastText uses a continuous bag-of-words model, which takes into account a symmetric window of words surrounding the word in question and maximizes the log-likelihood of the probability of the word given its context. Calculating the cosine-similarity between word vectors from fastText gives us a score of how similar the contexts of these words are. Therefore, words that occur in the same context (high lexical co-occurrence) would have a larger cosine-similarity score.

It must be noted that fastText supplements the lexical co-occurrence information by simultaneously training weight vectors for each position relative to the word in question, as well as adding sub-word information (bag-of-character n-gram vectors for each character forming the words). The similarity judgment prediction given by fastText may be more higher than a model using purely lexical co-occurrence information. However, models that use only lexical co-occurrence information (with or without a simple position weight system that gives more weight to words that are closer to the word in question) also perform well in similarity prediction tasks. For this study, due to the lack of large cleaned corpora for Mandarin and Vietnamese, we leverage fastText pre-trained models on the languages we are interested in @Grave2018.-->

<!-- condense to 2 sentences --> Systems like fastText and word2vec have been demonstrated to be good predictors for similarity judgments. @Liu2019 showed that fastText performed well as the data for a system matching responses in a knowledge base by how similar they are to a customer’s query. A study by @Jatnika2019 tested a word2vec model trained on the English wikipedia corpus against 353 word pairs labeled with similarity values based on human judgment. The word2vec model correlates moderately with human judgment. fastText has also been shown to be sensitive towards cultural effects on word meanings: @Thompson2020 applied fastText to language-specific Wikipedia corpus (among others) to generate a “semantic neighborhood” of 1010 meanings in different languages. The study showed that languages that are spoken by more similar cultures, more geographically proximate and/or historically related had more semantic neighborhoods that aligned. This is a proof of concept for our use of fastText in different languages with varying levels of relatedness.

@@

<!-- Additionally, outside of language, cross-cultural differences could come from a couple different mechanisms:  -->
@Ji2004 suggest that the notion of similarity is what varies across cultural groups, but it is possible that similarity is regarded in the same way and it is actually the input to this similarity judgment that varies across cultural contexts. It may be possible to gain traction on potential mechanisms by examining whether variation in similarity judgments co-vary with environmental statistics that differ across cultural and linguistic contexts.

In this study, we ask whether variation in lexical co-occurrence can account for the cross-context (culture and language) differences observed in how people evaluate similarity. First, we attempt to replicate @Ji2004 by running a study comparing English speakers in the US and Mandarin speakers in mainland China. Second, we evaluate whether the differences observed between English and Mandarin speakers’ similarity judgments also extend to comparisons between English and Vietnamese, by running a similar study between English speakers in the US and Vietnamese speakers in Vietnam. Vietnam is a Southeast Asian country that borders China and is historically greatly influenced by Chinese culture [@Hui2002]. Therefore, it serves as a suitable cultural context to investigate whether the claim made by @Ji2004 and previous studies, that Eastern and Western cultures have different notions of similarity, extends beyond mainland China.

We then ask to what extent similarity judgments show cross-context differences that align with co-occurrence patterns in participants’ languages. We would only see a main effect of cultural context if differing notions of similarity drive these cross-cultural differences. Meanwhile, if we do see variation in responding trial-to-trial that tracks with lexical co-occurrence statistics (as a rough proxy combining both linguistic and cultural context), this suggests that cross-cultural notions of similarity might be similar, and it is environmental statistics that guide the differences in similarity judgments.

To preview our results, we find that, while we do not see overall group differences (between the East Asian contexts and the US) in similarity, we do find language-specific co-occurrence can indeed explain cross-context differences in similarity judgments. In contrast to previous accounts, by which culture induces differing conceptions of similarity (varying between taxonomic and thematic), these findings provide an alternative, and more specific, account by which language may explain these cross-cultural differences without invoking variation in notions of similarity.

There are several important limitations of our approach. While we discuss cross-cultural variability at the level of countries or larger world areas, these are not cultural monoliths. For convenience, we operationalize culture at the level of country, based on where participants were raised. It is an open question whether performance in our participant populations (of relatively young and well-educated adults) is representative of the broader country. This is especially true for societies with substantial ethnic and cultural variation such as the US. We expect that our data is likely to underestimate variation both within and between the countries we sample from.

Language, culture, cognition, and individual experiences are intertwined in complex causal relationships: cultural features (farming) can lead to differences in individuals’ experiences (seeing cows and grass) and in turn these can influence language (talking more about cows and grass). But cultural features (a thematic orientation) could also more directly cause individuals to talk differently about the same experiences (mentioning what cows eat rather than what other animals cows are like). In this study, we measure language and its relation to cross-cultural differences in categorization, but these relations test only the plausibility of a language-based account; they cannot establish the direction of causality.
